Open-source tools in R for landscape ecology

### 2021-04-26 Comments_Reviewer.Rmd
Addressing the last final comments

@@

@@

*Reference:* Rayfield B, MJ Fortin, A Fall. 2011. Connectivity for conservation: a framework to classify network measures. Ecology 92(4): 847-858.

@@

**Reply**: Thank you for the general encouraging words about the manuscript. 
We agree that a definition of landscape ecology is not needed given the target audience of the journal. 
Thus, we shortened the introduction. 
\textcolor{red}{Maybe move section of powerful R to introduction? But, comprehensive comparison with other software outside of scope} \vskip 0.5cm

@@

**Reply**: All authors of the manuscript are experienced R users with many years of experience using the programming language. 
Additionally, all authors even contributed several packages to the community themselves. 
Given the different background of all authors, collectively probably most packages were used by at least one of the authors, but probably not all. 
Nevertheless, we also believe that this is not necessarily important for this review. 
Furthermore, we agree with the reviewer that more specific recommendation which package to use could be useful, however, we also think that this is very difficult. 
Reasons for this include, besides others, personal preferences of the users or analysis specific circumstances. 
Lastly, the review should be as neutral as possible without favoring one package over another because pros and cons of package differ between users and all were developed with a specific idea in mind.
\textcolor{red}{Refer to new how to get started box} \vskip 0.5cm

@@

@@

@@

@@

**Reply**: We added some explanation that raster and terra already provide solutions to this issue by processing only chunks that can be stored in memory. 
Additionally, we added some information about parallel processing as another possible solution to these issues. 
However, we believe that discussing underlying issues of R memory management is outside the scope if this paper. \vskip 0.5cm

@@

@@

**Reply**: Thank you very much for the kind and encouraging words about the manuscript.
We are happy to hear that the reviewer thinks the manuscript might be useful for ecologists that would like to analyze data through R. \vskip 0.5cm

>>>>

**Reply**: Thank you very much for considering the manuscript for publication. 
We appreciate the reviewers comments and think they all help to improve the manuscript.
We agree with reviewer 1 that the definition of landscape ecology could be shortened given the target audience of the journal. 
Thus, the introduction now puts a stronger focus on advantages of open-source software and especially R. 
Furthermore, we added explanations for all packages that were missing in the original submission as suggested by reviewer 1.
However, we think that given recommendations which packages to use is very difficult for several reasons. 
First, packages have different advantages and disadvantages that differ depending on e.g., the research question, the available data, or the personal preferences of the users.
Second, we believe that trying out several packages for similar task can actually beneficial and strengthen trust in the method.
But, by modifying Figure 1 and adding Box 1 we hope to give some more guidelines on how to organize an analysis.
We agree that the survey is not the core of the paper, nevertheless, the comments by both reviewers are very valuable and improve the conclusions we can draw from it. 
Thus, we added some explanation about the survey reach and what this means for the results of the survey. \vskip 0.5cm

@@

**Reply**: We agree that the of Rayfield et al. 2011 is great, but we also want to point out that it was developed for a more specific topic (i.e., connectivity), while this manuscript deals with the whole field of landscape ecology.
Thus, we believe that a similar typology for all packages included in this manuscript would be too complicated to be of actual help.
Thank you for this great idea to add boxes. 
We added a box "Where to start" to give some major steps how to approach landscape ecology in R for novice users. 
We hope this will also at least partly address the issue raised by reviewer 1 about recommendations.
We also thought about a second box for mode advanced users and even developers, but decided that the target audience of the paper are most likely novice to intermediate R users and not developers and the topic of developing own packages specifically for the landscape ecology community could maybe be a future manuscript.
Also, we added a short explanation of the tidyverse and its strength.
Lastly, we provide more information about the annotated references.
\vskip 0.5cm

@@

**Reply**: Again, thank you very much for the appreciated and valuable suggestions for improving the manuscript and we hope that the revised manuscript improved compared to the previous version with help of the reviewers. 
All changes to the manuscript are indicated by the blue color. \vskip 0.5cm

@@

**Reply**: Thank you for the general encouraging words about the manuscript. 
We agree that such a long definition of landscape ecology is not needed given the target audience of the journal. 
Thus, we shortened the introduction related to landscape ecology in general and rearranged other sections of the introduction.
Most of the introduction now introduces the advantages and strength of open-source software and in particular R. While all of the authors are enthusiastic R users, we think that a comprehensive discussion of R vs. other open-source software is outside of the scope of this paper and more important not expedient in general since other software might work better for other people (e.g., Python). \vskip 0.5cm

@@

**Reply**: All authors of the manuscript are experienced R users with many years of experience using the programming language. 
Additionally, all authors even contributed several packages to the community themselves. 
Given the different background of all authors, collectively probably most packages were used by at least one of the authors, but probably not all. 
Nevertheless, we also believe that this is not necessarily important for this review. 
Furthermore, we agree with the reviewer that more specific recommendation which package to use could be useful, however, we also think that this is very difficult. 
Reasons for this include, besides others, personal preferences of the users or analysis specific circumstances. 
Lastly, the review should be as neutral as possible without favoring one package over another because pros and cons of package differ between users and all were developed with a specific idea in mind.
But, we think that our new "Box 1: Where to start" hopefully can assist novice users with their workflow. \vskip 0.5cm

@@

**Reply**: We hope that the modified Figure 1, the added Box 1 and some added explanation in the conclusions explain how using R can facilitate a reproducible workflow in landcape ecology. \vskip 0.5cm

@@

**Reply**: While there are certainly many ways to define landscape ecology, we agree that pattern is a central term and included it in the introduction. \vskip 0.5cm

@@

**Reply**: Thank you for this valuable remark. We tried to add additional explanations to concepts which might be underexplained to especially novice users which hopefully increases the value of the manuscript. \vskip 0.5cm

@@

**Reply**: We added some explanation that raster and terra already provide solutions to this issue by processing only chunks that can be stored in memory. 
Additionally, we added some information about parallel processing as another possible solution to these issues. 
However, we believe that discussing underlying issues of R's memory management is outside the scope if this paper. \vskip 0.5cm

@@

**Reply**: We are aware that there a many packages listed in this manuscript which might be overwhelming, especially given that some packages cover similar tasks. 
We hope that especially Box 1, but also the restructured Figure 1 help landscape ecologists to find suitable packages. 
Furthermore, we think that the structure in Table 1 actually helps to find packages related to specific task. 
Lastly, we think that trying out similar packages for the same task can actually be beneficial, as also added to Box 1. \vskip 0.5cm

@@

**Reply**: Thank you very much for the kind and encouraging words about the manuscript.
We are happy to hear that the reviewer thinks the manuscript might be useful for landscape ecologists that would like to analyze data through R. 
We hope by including the appreciated and valuable comments by two reviewers we even increased the usefulness of the manuscript. \vskip 0.5cm

### 2021-04-26 paper/paper.Rmd
Addressing the last final comments

## A short introduction to landscape ecology {#sec:landscape_ecology}

While human activities have altered the landscapes for millennia \cite{Ellis2011,Ellis2015}, in the recent centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important for science, conservation, or management \cite{With2019}.

Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
In this context, landscape ecology predominantly focuses on i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchanges within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) and can guide how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize spatial data.

## Open-source software and *R* {#sec:open_source}

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}. 
\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g. Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

The *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).
The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
Thus, in the first part of this article, we present an overview of existing *R* packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are presently missing for landscape ecology.

@@

A comprehensive collection of *R* software packages exists to handle the most common tasks of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, *R* allows for transparent and reproducible workflows.
\textcolor{blue}{This script-based characteristic of the *R* programming language allows to easily share and reproduce analysis, as demonstrated by the R script in the appendix, which is a big advantage over point-and-click software interfaces.}
This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

@@

\noindent \textbullet \textbullet \cite{Bivand2020}: Overview of general developments of *R* to analyze spatial data.

\noindent \textbullet \textbullet \cite{Lovelace2019}: Textbook about spatial data handling with applied examples in the *R* programming language.

\noindent \textbullet \cite{Lai2019}: Review of the *R* programming language in ecology.

\noindent \textbullet \cite{Gustafson2019}: Review of quantifying landscape spatial pattern.

>>>>

Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
Thus, landscape ecology emphasizes \textcolor{blue}{spatial} patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these \textcolor{blue}{spatial patterns}.

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}. 
\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
Thus, the *R* programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the *R* environment (Fig. \@ref(fig:fig-workflow)).

The growing body of *R* packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the *R* programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
Thus, in the first part of this article, we present an overview of existing *R* packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are presently missing for landscape ecology.

@@

A comprehensive collection of *R* software packages exists to handle the most common tasks of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same  \textcolor{blue}{scripted} programming environment, *R* allows for transparent and reproducible workflows.
\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the *R* programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the *R* script in the appendix.
This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore are often only available for certain operating systems.}
This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

@@

\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of *R* to analyze spatial data demonstrating the capabilities of the programming language to anlyze spatial data.}

\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Textbook about spatial data handling and analyzing covering all common tasks with applied and helpful code examples in the *R* programming language.}

\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state-of-the-art for quantifying landscape spatial pattern including lessons learned by the author.}

\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the *R* programming language is wideley used in ecology including possible reasons for the wide spread use.}

### 2021-04-26 paper/paper.tex
Addressing the last final comments

\hypertarget{sec:landscape_ecology}{%
\subsection{A short introduction to landscape ecology}\label{sec:landscape_ecology}}

While human activities have altered the landscapes for millennia \cite{Ellis2011,Ellis2015}, in the recent centuries, the effects of humans on landscapes have increased to an unknown high, known as the Anthropocene \cite{Crutzen2002}.
Today, almost all landscapes are directly or indirectly influenced by human activities \cite{Vitousek1997}.
Thus, understanding the complex interactions between landscapes and ecological processes becomes increasingly important for science, conservation, or management \cite{With2019}.

Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
In this context, landscape ecology predominantly focuses on i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes, and exchanges within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly, iv) and can guide how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize spatial data.

\hypertarget{sec:open_source}{%
\subsection{\texorpdfstring{Open-source software and \emph{R}}{Open-source software and R}}\label{sec:open_source}}

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.

One successful example of an open-source project is the \emph{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of \emph{R} to handle spatial data in general can be found in \cite{Bivand2020}.

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

The \emph{R} programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig. \ref{fig:fig-workflow}).
The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
Thus, in the first part of this article, we present an overview of existing \emph{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are presently missing for landscape ecology.

@@

A comprehensive collection of \emph{R} software packages exists to handle the most common tasks of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same \textcolor{blue}{scripted} programming environment, \emph{R} allows for transparent and reproducible workflows.
\textcolor{blue}{This script-based characteristic of the *R* programming language allows to easily share and reproduce analysis, as demonstrated by the R script in the appendix, which is a big advantage over point-and-click software interfaces.}
This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

@@

\noindent \textbullet \textbullet \cite{Bivand2020}: Overview of general developments of \emph{R} to analyze spatial data.

\noindent \textbullet \textbullet \cite{Lovelace2019}: Textbook about spatial data handling with applied examples in the \emph{R} programming language.

\noindent \textbullet \cite{Lai2019}: Review of the \emph{R} programming language in ecology.

\noindent \textbullet \cite{Gustafson2019}: Review of quantifying landscape spatial pattern.

>>>>

Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes they occur in and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
Thus, landscape ecology emphasizes \textcolor{blue}{spatial} patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these \textcolor{blue}{spatial patterns}.

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratizes scientific research as free-of-cost software removes one gatekeeper for researchers without access to proprietary software.
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

One successful example of an open-source project is the \emph{R} programming language, and its extensions called packages \cite{RCoreTeam2019}.
\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the *R* programming language \cite{Wickham2015}}.
Firstly released in 1995 \cite{Smith2016}, today \textcolor{blue}{this} programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006,Lovelace2019}.
Thus, the \emph{R} programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig. \ref{fig:fig-workflow}).

The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: \emph{Analysis of Spatial Data} \cite{Bivand2019a} and \emph{Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely higher.
The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of \emph{R} to handle spatial data in general can be found in \cite{Bivand2020}.

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
\textcolor{blue}{We acknowledges that there are reasons and circumstances to also use other software tools and progamming languages, however, a comprehensive discussion of pro and cons is outside the scope of this manuscript.}
Thus, in the first part of this article, we present an overview of existing \emph{R} packages for landscape ecology and closely related fields (Table \ref{tab:packages}).
\textcolor{blue}{We included only packages that were available on the official *R* package archive network CRAN at the time the manuscript was prepared. 
This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
In the second part, we present a survey in which we asked the community how they currently use the \emph{R} programming language and to identify topics for which \emph{R} packages are presently missing for landscape ecology.

@@

A comprehensive collection of \emph{R} software packages exists to handle the most common tasks of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same \textcolor{blue}{scripted} programming environment, \emph{R} allows for transparent and reproducible workflows.
\textcolor{blue}{Because scripts including all analysis steps and parameter settings can easily be shared, the script-based characteristic of the *R* programming language allows to easily share and reproduce analysis across all major operation systems (e.g., Windows, macOS, Linux), as demonstrated by the *R* script in the appendix.
This is a big advantage over point-and-click software interfaces where analysis steps and parameter settings can only be described but not shared, and furthermore are often only available for certain operating systems.}
This also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

@@

\noindent \textbullet \textbullet \cite{Bivand2020}: \textcolor{blue}{Recent and comprehensive overview of general developments of *R* to analyze spatial data demonstrating the capabilities of the programming language to anlyze spatial data.}

\noindent \textbullet \textbullet \cite{Lovelace2019}: \textcolor{blue}{Textbook about spatial data handling and analyzing covering all common tasks with applied and helpful code examples in the *R* programming language.}

\noindent \textbullet \cite{Gustafson2019}: \textcolor{blue}{This review presents an overview of the developments and the state-of-the-art for quantifying landscape spatial pattern including lessons learned by the author.}

\noindent \textbullet \cite{Lai2019}: \textcolor{blue}{This review demonstrates that the *R* programming language is wideley used in ecology including possible reasons for the wide spread use.}
